An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-11-27 paper/2.minbigul.tex
small thing
fix indent + scaled tick

minBiGUL, our target language in this paper, is a subset of BiGUL which is a simple yet powerful putback-based bidirectional language. BiGUL supports two transformations: a forward transformation $get$ producing a view from a source and a backward transformation $put$ taking a source and a modified view to produce an updated source. Intuitively, if we have a BiGUL program $bx$, these two transformations are following functions:\\
    $\tab \getbxinline{bx} : s \to v,
    \tab \putbxinline{bx} : s * v \to v$

BiGUL is well-behaved [?] since two functions $\putbxinline{bx}$ and $\getbxinline{bx}$ satisfy the round-trip laws as follows:\\
    $\tab \putbx{bx}{s}{(\getbx{bx}{s})} = s \qquad [\textsc{GetPut}]\\
    \tab \getbx{bx}{(\putbx{bx}{s}{v})} = v \qquad [\textsc{PutGet}]$

The \textsc{GetPut} \cite{putget} law means that if there is no change to the view, there should be no change to the source. The \textsc{PutGet} \cite{putget} law means that we can recover the modified view by applying the forward transformation to the updated source.

minBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which are satisfy two above laws. In addition, we restrict adaptive cases of BiGUL on minBiGUL. Then $put$ and $get$ satisfy one more law, \textsc{PutPut} \cite{putput}, like the following:\\
    $\tab \putbx{bx}{(\putbx{bx}{s}{v'})}{v} = \putbx{bx}{s}{v} \qquad [\textsc{PutPut}]$

@@

$\tab bx := Skip \ h \
        | \ Replace \
        | \ Prod \ bx_1 \ bx_2 \
        | \ RearrS \ f_1 \ f_2 \ bx \
        | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtab | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \
        | \ Compose \ bx_1 \ bx_2$

@@

        $\putbx{Skip \ h}{s}{v} = \\
            \tab \sif{h \ s = v}{s}{\text{fail}}$

        $\putbx{Replace}{s}{v} = v$

        $\putbx{\product{bx_1}{bx_2}}{(s_1, s_2)}{(v_1, v_2)} = \\
            \tab ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}))$

        $\putbx{\rearrs{f_1}{f_2}{bx}}{s}{v} = \\
            \tab f_2 \ (\putbx{bx}{(f_1 \ s)}{v})$

        $\putbx{\rearrv{g_1}{g_2}{bx}}{s}{v} = \\
            \tab \putbx{bx}{s}{(g_1 \ v)}$

        $\putbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} = \\
            \tab \text{if} \ {cond_{sv} \ s \ v} \\
            \tab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \\
            \tab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \\
            \tab \text{fi} \ cond_{s} \ s'; \ \text{return} \ s'$

        $\putbx{bx_1 \circ bx_2}{s}{v} = \\
            \tab \putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}$
    \end{definition}
\columnbreak
    \begin{definition}
        $\getbx{bx}{s}$

        $\getbx{Skip \ h}{s} = \\ 
            \tab h \ s$

        $\getbx{Replace}{s} = s$

        $\getbx{\product{bx_1}{bx_2}}{(s_1,s_2)} = \\
            \tab ((\getbx{bx_1}{s_1}),(\getbx{bx_2}{s_2}))$

        $\getbx{\rearrs{f_1}{f_2}{bx}}{s} = \\ 
            \tab \getbx{bx}{(f_1 \ s)}$

        $\getbx{\rearrv{g_1}{g_2}{bx}}{s} = \\ 
            \tab g_2 \ (\getbx{bx}{s})$

        $\getbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s} = \\
            \tab \text{if} \ {cond_{s} \ s} \\
            \tab \text{then} \ v' \Leftarrow \getbx{bx_1}{s} \\ 
            \tab \text{else} \ v' \Leftarrow \getbx{bx_2}{s} \\ 
            \tab \text{fi} \ {cond_{sv} \ s \ v'}; \ \text{return} \ v'$

        $\getbx{bx_1 \circ bx_2}{s} = \\ 
            \tab \getbx{bx_2}{(\getbx{bx_1}{s})}$
    \end{definition}
\end{multicols}

We define the semantics of $put$ and $get$ as in definitions 1 and 2 respectively. Instead of using $v'$ in the $put$ direction like figures \ref{fig:eval-phead}, \ref{fig:eval-comp-phead} and \ref{fig:eval-comp-phead-2}, we just simply use notation $v$ as the updated view. The later definitions also use this convention. \textcolor{red}{$Leftarrow$ means assignments.}

In the above definitions, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, i.e. $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false. If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tab \equiv \text{if }E_1 = true \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else assert } false; \}\\
    \qtab \text{else } \{ X_2; \text{ if } E_2 = false \text{ then } S \text{ else assert } false; \}$

% Next, let's take a look at some examples to better understand about minBiGUL. We start with quite obvious things as follows:

% $\tab \putbx{Skip \ (\lambda x.(x*x))}{10}{100} = 10 \qtab \ \getbx{Skip \ (\lambda x.(x*x))}{10} = 100\\
%     \tab \putbx{Skip \ (\lambda \_.())}{1}{()} = 1 \qtabs{3} \getbx{Skip \ (\lambda \_.())}{1} = ()\\
%     \tab \putbx{Replace}{1}{100} = 100 \qtabs{3} \getbx{Replace}{1} = 1
% $

We now consider the definition of $phead$ in minBiGUL:

$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$

The above program rearranges the source, a non-empty list, to a pair of its head element $s$ and its tail $ss$, and the view to a pair $(v, ())$, then we can use $v$ to replace $s$ and $()$ to keep $ss$. Intuitively, $\putbx{phead}{s_0}{v_0}$ returns a list whose head is $v_0$ and tail is the tail of $s_0$, and $\getbx{phead}{s_0}$ returns the head of the list $s_0$. For instance, $\putbx{phead}{[1,2,3]}{100} = [100,2,3]$ and $\getbx{phead}{[1,2,3]} = 1$. If we wanna update the head element of the head element of a list of lists by using the view, we can define a composition like $phead \circ phead$. For example:\\
\tab $\putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$\\
\tab $\getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$

>>>>

minBiGUL, our target language in this paper, is a subset of BiGUL which is a simple yet powerful putback-based bidirectional language. BiGUL supports two transformations: a forward transformation $get$ producing a view from a source and a backward transformation $put$ taking a source and a modified view to produce an updated source. Intuitively, if we have a BiGUL program $bx$, these two transformations are following functions:

    $\qtab \getbxinline{bx} : s \to v, \quad \putbxinline{bx} : s * v \to v$

BiGUL is well-behaved [?] since two functions $\putbxinline{bx}$ and $\getbxinline{bx}$ satisfy the round-trip laws as follows:

    $\qtab \putbx{bx}{s}{(\getbx{bx}{s})} = s \qquad [\textsc{GetPut}]$

    $\qtab \getbx{bx}{(\putbx{bx}{s}{v})} = v \qquad [\textsc{PutGet}]$

The \textsc{GetPut} \cite{Pacheco:2014:MCP:2543728.2543737} law means that if there is no change to the view, there should be no change to the source. The \textsc{PutGet} \cite{Pacheco:2014:MCP:2543728.2543737} law means that we can recover the modified view by applying the forward transformation to the updated source.

minBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which satisfy two above laws. In addition, we restrict adaptive cases of BiGUL on minBiGUL. Then $put$ and $get$ satisfy one more law, \textsc{PutPut} \cite{putput}, like the following:

    $\qtab \putbx{bx}{(\putbx{bx}{s}{v'})}{v} = \putbx{bx}{s}{v} \qquad [\textsc{PutPut}]$

@@

    $bx := Skip \ h \
        | \ Replace \
        | \ Prod \ bx_1 \ bx_2 \
        | \ RearrS \ f_1 \ f_2 \ bx \
        | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtabs{2} | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \
        | \ Compose \ bx_1 \ bx_2$

@@

        \noindent $\putbx{Skip \ h}{s}{v} = \\
            \tab \sif{h \ s = v}{s}{\text{fail}}$

        \noindent $\putbx{Replace}{s}{v} = v$

        \noindent $\putbx{\product{bx_1}{bx_2}}{(s_1, s_2)}{(v_1, v_2)} = \\
            \tab ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}))$

        \noindent $\putbx{\rearrs{f_1}{f_2}{bx}}{s}{v} = \\
            \tab f_2 \ (\putbx{bx}{(f_1 \ s)}{v})$

        \noindent $\putbx{\rearrv{g_1}{g_2}{bx}}{s}{v} = \\
            \tab \putbx{bx}{s}{(g_1 \ v)}$

        \noindent $\putbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} =\\
            \tab \text{if} \ {cond_{sv} \ s \ v} \\
            \tab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \\
            \tab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \\
            \tab \text{fi} \ cond_{s} \ s'; \ \text{return} \ s'$

        \noindent $\putbx{bx_1 \circ bx_2}{s}{v} = \\
            \tab \putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}$
    \end{definition}
\columnbreak
    \begin{definition}
        $\getbx{bx}{s}$

        \noindent $\getbx{Skip \ h}{s} = \\ 
            \tab h \ s$

        \noindent $\getbx{Replace}{s} = s$

        \noindent $\getbx{\product{bx_1}{bx_2}}{(s_1,s_2)} = \\
            \tab ((\getbx{bx_1}{s_1}),(\getbx{bx_2}{s_2}))$

        \noindent $\getbx{\rearrs{f_1}{f_2}{bx}}{s} = \\ 
            \tab \getbx{bx}{(f_1 \ s)}$

        \noindent $\getbx{\rearrv{g_1}{g_2}{bx}}{s} = \\ 
            \tab g_2 \ (\getbx{bx}{s})$

        \noindent $\getbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s} = \\
            \tab \text{if} \ {cond_{s} \ s} \\
            \tab \text{then} \ v' \Leftarrow \getbx{bx_1}{s} \\ 
            \tab \text{else} \ v' \Leftarrow \getbx{bx_2}{s} \\ 
            \tab \text{fi} \ {cond_{sv} \ s \ v'}; \ \text{return} \ v'$

        \noindent $\getbx{bx_1 \circ bx_2}{s} = \\ 
            \tab \getbx{bx_2}{(\getbx{bx_1}{s})}$
    \end{definition}
\end{multicols}

We define the semantics of $put$ and $get$ as in definitions 1 and 2 respectively. Instead of using $v'$ in the $put$ direction like figures \ref{fig:eval-phead}, \ref{fig:eval-comp-phead} and \ref{fig:eval-comp-phead-2}, we just simply use notation $v$ as the updated view. The later definitions also follow this convention.

In the above definitions, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$ where $\Leftarrow$ means assignments. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, i.e. $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false. If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tabs{2} \ \equiv \text{if }E_1 = true \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else assert } false; \}\\
    \tabs{4} \text{else } \{ X_2; \text{ if } E_2 = false \text{ then } S \text{ else assert } false; \}$

\noindent We now consider the definition of $phead$ in minBiGUL:\\
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$

The above program rearranges the source, a non-empty list, to a pair of its head element $s$ and its tail $ss$, and the view to a pair $(v, ())$, then we can use $v$ to replace $s$ and $()$ to keep $ss$. Intuitively, $\putbx{phead}{s_0}{v_0}$ returns a list whose head is $v_0$ and tail is the tail of $s_0$, and $\getbx{phead}{s_0}$ returns the head of the list $s_0$. For instance, $\putbx{phead}{[1,2,3]}{100} = [100,2,3]$ and $\getbx{phead}{[1,2,3]} = 1$. If we wanna update the head element of the head element of a list of lists by using the view, we can define a composition like $phead \circ phead$. For example:

$\putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$

$\getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$
